---
title: "Quantitative Data Analysis Assignment 1"
description: |
  Assignment 1 for DACSS 603 course 'Quantitative Data Analysis': "Descriptive Statistics, Probability, Statistical Inference, and Comparing Two Means"
categories:
  - statistics
  - homework
author:
  - name: Kristina Becvar
    url: https://www.kristinabecvar.com
slug: kbec2022qda1
date: 02-23-2022
base_url: https://www.kristinabecvar.com
output:
  distill::distill_article:
    self_contained: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(dplyr)
library(ggplot2)

```

## Question 1

### Surgical Procedure - Representative Sample

Surgical Procedure | Sample Size | Mean Wait Time | Standard Deviation
------------- | ------------- | ------------- | -------------
Bypass | 539 | 19 | 10
Angiography | 847 | 18 | 9

* Construct the 90% confidence interval to estimate the actual mean wait time for each of the two procedures. 

* Is the confidence interval narrower for angiography or bypass surgery?

## Answer 1

I calculated the answer by first calculating the standard error for each procedure given the mean, standard deviation, and sample size for each. I do so using 0.95 for the qnorm function so that I can determine the 5% confidence level for both the right and left side of the normal distribution, since the sample is larger than n=30. By calculating the 5% margin for each side of the distribution, this gives me the effective 90% confidence interval overall.

```{r echo=TRUE}

#Calculate the actual mean wait time for the bypass:

xbar1 <- 19 #sample mean
sd1a <- 10 #sample standard deviation
n1a <- 539 #sample size

error1a <- qnorm(0.95)*sd1a/sqrt(n1a)
error1a

lower1a <- xbar1-error1a
upper1a <- xbar1+error1a

lower1a
upper1a

```

```{r echo=TRUE}

#Calculate the actual mean wait time for the angiography:

xbar1b <- 18 #sample mean
sd1b <- 9 #sample standard deviation
n1b <- 847 #sample size

error1b <- qnorm(0.95)*sd1b/sqrt(n1b)
error1b

lower1b <- xbar1b-error1b
upper1b <- xbar1b+error1b

lower1b
upper1b

```

Next, I created a data frame with the information.

```{r code_folding=TRUE}

#Create a data frame with the information:

df1 <- data.frame( c('Bypass', 'Angiography')
                   ,c(19, 18)
                   ,c(539, 847)
                   ,c(10, 9)
                   ,c(18.29151, 17.49134)
                   ,c(19.70849, 18.50866))
names(df1) <- c('Procedure', 'Mean', 'Sample', 'SD', 'Lower', 'Upper')

df1

```

Finally, I created a plot to visualize the results. The visualization communicates that for each procedure, there is a range of values where we can expect 90% of the estimates to include the population mean given the sample mean and standard deviation.

```{r code_folding=TRUE}

gg1 <- ggplot(data = df1)
gg1 <- gg1 + geom_point(aes(x = Procedure, y = Mean), size = 5, color = "blue")
gg1 <- gg1 + geom_errorbar(aes(x = Procedure, y = Mean, ymin=Lower, ymax=Upper), width=.1, color = "blue")
gg1 <- gg1 + geom_text(aes(x = Procedure, y = Lower, label = round(Lower, 2)), hjust = -1)
gg1 <- gg1 + geom_text(aes(x = Procedure, y = Upper, label = round(Upper, 2)), hjust = -1)
gg1 <- gg1 + geom_text(aes(x = Procedure, y = Mean, label = round(Mean, 2)), hjust = -0.5)
gg1 <- gg1 + labs(x = "Procedure", y = "Mean")
gg1 <- gg1 + labs(title = "Chart Showing Mean With Upper and Lower Confidence Intervals at 90%")
gg1 <- gg1 + theme_classic()

gg1


```

* For the angiography, we can be 90% sure that the population mean for the wait time to the procedure falls between 17.49 and 18.51 days.

* For the bypass, we can be 90% sure that the population mean for the wait time to the procedure falls between 18.29 and 19.71 days.

* The confidence interval was narrower for the angiography surgery.

## Question 2

A survey of 1031 adult Americans was carried out by the National Center for Public Policy. Assume that the sample is representative of adult Americans. Among those surveyed, 567 believed that college education is essential for success. 

* Find the point estimate, p, of the proportion of all adult Americans who believe that a college education is essential for success. 

* Construct and interpret a 95% confidence interval for p.

## Answer 2

To construct the 95% confidence interval, I began by calculating the point sample estimate of the population proportion:

```{r echo=TRUE}

#Calculate the point sample estimate of the population proportion using (p = x/n)

x2 = 567 #affirmative response size
n2 = 1031 #sample size - survey participants

p2 <- x2/n2
p2

```

* This tells me that the sample proportion of those who believe a college education is essential for success is ~45%.

Since np >= 5 and n(1-p) >= 5, I know that I will calculate the confidence interval of that population proportion as follows: p +/- z * square root of (p) x (1-p)/n

```{r echo=TRUE}

#Calculate the confidence interval for p:

error2 <-qnorm(0.975)*sqrt(p2*(1-p2)/n2)
error2

lower2 <- p2-error2
upper2 <- p2+error2

lower2
upper2


```

* This tells me that we can be 95% confident that the proportion of adult Americans who believe that a college education is essential for success lies between 51.96% and 58.03% of the population.

Alternatively, using the R prop.test function, I can compare the calculation to my manual calculation and find it is the same for finding the point, but slightly different (at 4 decimal points) on the calculation of the confidence interval.

``` {r echo=TRUE}

prop.test(x2, n2)

```

## Question 3

Suppose that the financial aid office of UMass Amherst seeks to estimate the mean cost of textbooks per quarter for students. The estimate will be useful if it is within $5 of the true population mean (i.e. they want the confidence interval to have a length of $10 or less). The financial aid office is pretty sure that the amount spent on books varies widely, with most values between $30 and $200. They think that the population standard deviation is about a quarter of this range. 

* Assuming the significance level to be 5%, what should be the size of the sample?

## Answer 3

I will start by taking the information given and calculating the variables I need to know.

If the aid office believes the amount spent on books is between $30 and $200, I have a range of $170 to consider.

I want to be 95% confident that the interval estimate contains the population mean, so my z = 1.96.

I also know that the margin of error should be no more than +/- $5 on each end of the estimate, so my margin of error = 5

```{r echo=TRUE}

error3 <- (5)

#I need to calculate the standard deviation at the given estimate that it is a quarter of the range:

sd3 <- 170*0.25 #range * 25%
sd3 #standard deviation = 42.5

#Now I can calculate the sample size with the formula n=(zσ/M)2.

ss3 <- ((1.96*sd3)/error3)^2
ss3 #sample size

```

* Using these calculations, I can estimate that the financial aid office will need to use a sample size of 278 people.

## Question 4

According to a union agreement, the mean income for all senior-level workers in a large service company equals $500 per week. A representative of a women’s group decides to analyze whether the mean income μ for female employees matches this norm. For a random sample of nine female employees, ȳ = $410 and s = 90.

* A. Test whether the mean income of female employees differs from $500 per week. Include assumptions, hypotheses, test statistic, and P-value. Interpret the result.

* B. Report the P-value for Ha : μ < 500. Interpret.

* C. Report and interpret the P-value for H a: μ > 500.

(Hint: The P-values for the two possible one-sided tests must sum to 1.)

## Answer 4

* A. I will start by taking the information given and determining my hypotheses:

1. H0: The mean weekly earnings for the population of women at the company is μ=$500#

2. Ha: The mean weekly earnings for the population of women at the company is μ≠$500

I cannot assume that the population distribution is normal as I have a low sample size of 9 and it is not stated it is a random sample, but a representative sample.

My other assumptions are:

* population mean (mu4) = 500
* sample size (n4) = 9
* sample mean (xbar4) = 410
* sample standard deviation (sd4) = 90

I also need to make a decision about using a significance level of 5%.

I will use the test statistic formula to find the t-value: [t = (x̄) - (μ) / (sd/sqrt(n)]

```{r echo=TRUE}

a4 <- 500
n4 <- 9
xbar4 <- 410
sd4 <- 90

#Test statistic:

t4 <-(xbar4-a4)/(sd4/sqrt(n4))

t4

```

Given that my test statistic = (-3), I can determine the p-value is .00135*2 (to get the sum of both tail probabilities) or 0.0027.

**_Since  my confidence level is 0.05 and my p-value of 0.0027 < 0.05, I can reject the null hypothesis that the mean weekly earnings for the population of women at the company is μ=$500._**

```{r echo=TRUE}

#Then I take the t-statistic result (-3) and the degrees of freedom by taking "sample size - 1" or ("9" - 1) = 8.

pval4a <- pt(-3, 8)

pval4a

```

* B. For the alternative hypothesis Ha: μ < 500:

```{r echo=TRUE}

pval4b <- pt(-3, 8, lower.tail=FALSE)

pval4b

```
Per the hint, I can confirm that these are logical answers by adding the two probabilities together and confirming they equal 1:

```{r echo=TRUE}

pval4a+pval4b

```

## Question 5

Jones and Smith separately conduct studies to test H0: μ = 500 against Ha : μ ≠ 500, each with n = 1000. 

Jones gets ȳ = 519.5, with se = 10.0. Smith gets ȳ = 519.7.

* A. Show that t = 1.95 and P-value = 0.051 for Jones. Show that t = 1.97 and P-value = 0.049 for Smith.

* B. Using α = 0.05, for each study indicate whether the result is “statistically significant.”

* C. Using this example, explain the misleading aspects of reporting the result of a test as “P ≤ 0.05” versus “P > 0.05,” or as “reject H0” versus “Do not reject H0 ,” without reporting the actual P-value.

## Answer 5

* A. To show the t-scores, I will use the test statistic [t = (ybar) - (μ) / (se)] given the results for each:

```{r echo=TRUE}

#Test statistic for Jones:

ybar5a <- 519.5
n5 <- 1000
a5 <- 500
se5 <- 10.0

t5a <-(ybar5a-a5)/(se5)

t5a

#Test statistic for Smith:

ybar5b <- 519.7
n5 <- 1000
a5 <- 500
se5 <- 10.0

t5b <-(ybar5b-a5)/(se5)

t5b

```

* To show the p-values, I will use the pt() function and use the t-statistic results from each of the tests and the degrees of freedom by taking "sample size - 1" or ("100" - 1) = 999. I will need to multiply each result by 2 to account for the probabilities in each tail of the normal distribution.


```{r echo=TRUE}

#For Jones' results:

pval5a <- pt(1.95, 999, lower.tail = FALSE) * 2

pval5a

#For Smith's results:

pval5b <- pt(1.97, 999, lower.tail = FALSE) * 2

pval5b

```

* B. To use α = 0.05 and look at each result and whether it is "statistically significant", I can compare the p-values directly to the confidence level of 0.5. Jones' results gave a p-value of 0.5145, which is just __over__ the threshold of the confidence level given of 0.5. Smith's results gave a p-value of 0.4911, which is just __under__ the threshold of the confidence level given of 0.5.  

Hypothesis tests tell us that if the p-value < α, we reject H0 and if p-value ≥ α, we do not reject H0. Given this general statistical guidance, only Smith's results would be considered "statistically significant". 

* C. The results in this example could be very misleading if only whether the results were reported as simply "rejecting" or "not rejecting" H0 or being "statistically significant" or not because that is leaving out vital information on how close the results were to the confidence level used. If the actual p-values were reported instead, they would reflect how marginally "significant" the results really were. The confidence is an artificial threshold that is subjectively applied, and in this case, the results were close enough that they should not be statistically reported as significally different. This is why we need to be sure we report the full p-values and confidence intervals.


## Question 6


Are the taxes on gasoline very high in the United States? According to the American Petroleum Institute, the per gallon federal tax that was levied on gasoline was 18.4 cents per gallon. However, state and local taxes vary over the same period. The sample data of gasoline taxes for 18 large cities is given below in the variable called gas_taxes.

gas_taxes <- c(51.27, 47.43, 38.89, 41.95, 28.61, 41.29, 52.19, 49.48, 35.02, 48.13, 39.28, 54.41, 41.66, 30.28, 18.49, 38.72, 33.41, 45.02)

* Is there enough evidence to conclude at a 95% confidence level that the average tax per gallon of gas in the US in 2005 was less than 45 cents? Explain.

## Answer

To answer whether there is enough evidence to conclude at a 95% confidence level that the average tax per gallon of gas in the US in 2005 was less than 45 cents, we need to use a left-tailed t-test. We will use the sample data in the variable "gas_taxes" in the t.test function. We know that the t.test() function uses the 95% confidence interval as a default, but it also uses a two-sided test as the default, so we need to provide the alternative argument "less", indicating a left-sided test.

```{r echo=TRUE}

gas_taxes <- c(51.27, 47.43, 38.89, 41.95, 28.61, 41.29, 52.19, 49.48, 35.02, 48.13, 39.28, 54.41, 41.66, 30.28, 18.49, 38.72, 33.41, 45.02)

t.test(gas_taxes, alternative = "less")

```

This t-test gives us a confidence interval of [inf - 44.67946]. Since the confidence interval includes amounts that are all __less__ than 45 cents, we have enough evidence to conclude that, at that confidence level, the average tax was __less__ than 45 cents.
