---
title: "DACSS 697E Assignment 3"
description: |
  Assignment 3 for DACSS 697E course 'Social and Political Network Analysis': "Grateful Research: Creating a Network"
categories:
  - networks
  - homework
  - grateful network
author:
  - name: Kristina Becvar
    url: https://www.kristinabecvar.com
slug: kbec2022networks2
date: 03-02-2022
base_url: https://www.kristinabecvar.com
output:
  distill::distill_article:
    self_contained: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

I am using a data set that is somewhat similar in structure to that of my final project to try and get a feel for the process of creating the appropriate network. After recovering data from thousands of New York Times articles pulled through their API on Afghanistan from a 2-year period, I will be analyzing the network of article authorship and themes of articles. To understand the process, I am using for my assignment a data set of co-writers of songs played by the Grateful Dead over their 30-year touring career that I compiled. While compiling the data, I added an attribute that represents the connections between co-writers as songs with the added observation of the number of times each song was played live. The nature of the band was that of a collaborative subculture where the energy of the live shows reflected the crowd's buy-in to the songs being played. Since the band was primarily one whose popularity was measured by ticket sales, not album sales.  I'm still not sure if that will serve appropriately as a 'weight' for the network data, so I need to explore the process more thoroughly.

```{r echo=FALSE, results=FALSE}

library(ggplot2) 
library(readr) 
library(igraph)
library(ggraph)
library(tidyverse)

suppressWarnings(expr)
options(ggrepel.max.overlaps = Inf)
```

Understanding this as an affiliation network, I first created a matrix linking actors (songwriters) to an event (songs). I began by assigning a unique ID to each actor and event. Taking the data I have pulled from my research, I created an affiliation spreadsheet with the songwriters as rows and the songs as columns. When a songwriter was affiliated with a song, there was a number in that matrix spot. However, I struggled to get this affiliation data into a network in R because I was using weights incorrectly, so I took a different approach.

In this example, I used a node list where unique IDs are numbers which correspond to the name of a songwriter. 

The edgelist is in a separate spreadsheet where the  first two columns are the IDs of the source and the target node (songwriter ID), regardless of whether the network is directed, for each edge. Each row contains an observation of a connection between writers for a given song, and since there are multiple collaborations, there may be multiple rows of writer combinations for a given song ID. If there was only one writer on a song, that songwriter's ID is indicated in both the source and target column for that song.

The following columns are edge attributes. In my edgelist, I have the two songwriters representing the co-writing relationship in columns "1" and "2", the song ID in column "3", the song name in column "4", and the number of times the corresponding song was played live is indicated in column "5".

I have NOT utilized the number of times the song was played live as a network weight at this point. Additionally, this edgelist format is not the ideal format, but it is the first step in the process I am working through to utilize different methods of working through the data. In the future, I will use the data in the form of an affiliation matrix.

```{r echo=TRUE}

# Loading nodes and vertices

gd_vertices <- read.csv("gd_nodes.csv")
gd_edgelist <- read.csv("gd_clean_data.csv")

```

Converting network data into igraph objects using the "graph.data.frame: function, which takes two data frames: d and vertices.  

"d" describes the edges of the network and "vertices" the nodes.

```{r echo=TRUE}

set.seed(1234)
grateful_data <- graph_from_data_frame(d = gd_edgelist, vertices = gd_vertices, directed = FALSE)

```

Now to check the vertices and edges in the graph I've created to ensure they represent the data accurately, and confirm that all of the attributes have been represented properly:

```{r echo=TRUE}

head(V(grateful_data)$name)
head(E(grateful_data)$song.id)
head(E(grateful_data)$song.name)
head(E(grateful_data)$weight)

is_directed(grateful_data)
is_weighted(grateful_data)
is_bipartite(grateful_data)

igraph::vertex_attr_names(grateful_data)
igraph::edge_attr_names(grateful_data)

```

Next I want to take a first look at the network:

```{r code_folding=TRUE}

plot(grateful_data)

```

It's basically plotting what I want it to illustrate, though I will need to do a lot more work to make the graph represent anything meaningful! 

Finishing the look at the basic network information such as the dyad and triad census: I have 558 mutual dyads and null value of "-233", with a warning that calling a dyad census on an undirected graph. This does indicate to me that the edgelist format is not the best representation of this data.

```{r code_folding=TRUE}

igraph::dyad.census(grateful_data)
igraph::triad.census(grateful_data)

```

Knowing this network has 26 vertices, I want to see if the triad census is working correctly by comparing the following data, which I can confirm using this calculation.

```{r code_folding=TRUE}

#possible triads in network
26*25*24/6

sum(igraph::triad.census(grateful_data))

```

Looking next at the global v. average local transitivity of the network:

```{r code_folding=TRUE}

#get global clustering cofficient: igraph
transitivity(grateful_data, type="global")
#get average local clustering coefficient: igraph
transitivity(grateful_data, type="average")

```

This transitivity tells me that the average network transitivity is significantly higher than the global transitivity, indicating, from my still naive network knowledge, that the overall network is generally more loose, and that there is a more connected sub-network.

Looking at the geodesic distance tells me that on average, the path length is just over 2.

```{r code_folding=TRUE}

average.path.length(grateful_data,directed=F)

```

Getting a look at the components of the network shows that there are 2 components in the network, and 25 of the 26 nodes make up the giant component with 1 isolate.

```{r code_folding=TRUE}

names(igraph::components(grateful_data))
igraph::components(grateful_data)$no 
igraph::components(grateful_data)$csize

```

This is a great start - now I can get to looking at the network density, centrality, and centralization.

The network density measure: First with just the call "graph.density" and then with adding "loops=TRUE". Since I'm using igraph, I know that its' default output assumes that loops are not included but does not remove them, which can be corrected with the addition of "loops=TRUE" per the course tutorials when comparing output to statnet. This gives me confidence that my network density is closer to 1.58.

```{r code_folding=TRUE}

graph.density(grateful_data)
graph.density(grateful_data, loops=TRUE)

```
The network degree measure: This gives me a clear output showing the degree of each particular node (songwriter). It is not suprising, knowing my subject matter, that Jerry Garcia is the highest degree node in this network as the practical and figurative head of the band. The other band members' degree measures are not necessarily what I expected, though. I did not anticipate that his songwriting partner, Robert Hunter, would have a lower degree than band members Phil Lesh and Bob Weir. Further, I did not anticipate that the degree measure of band member 'Pigpen' would be so high given his early death in the first years of the band's touring life.

```{r code_folding=TRUE}

igraph::degree(grateful_data)

```

To look further I will create a dataframe for easier review going forward.

```{r code_folding=TRUE}

grateful_nodes<-data.frame(name=V(grateful_data)$name, degree=igraph::degree(grateful_data))

grateful_nodes

```

A quick look at the summary statistics confirms for me the minimum, maximum, median, and mean node degree data.

```{r code_folding=TRUE}

summary(grateful_nodes)

```

Now I want to take a step back and try to visually represent this data better.

```{r code_folding=TRUE}

# Community detection algoritm 
community <- cluster_louvain(grateful_data) 

# Attach communities to relevant vertices
V(grateful_data)$color <- community$membership 

# Graph layout
layout <- layout.random(grateful_data) 

# igraph plot 
plot(grateful_data, layout = layout)

```

Better, but not quite.

```{r code_folding=TRUE}

ggraph(grateful_data, layout = "fr") +
  geom_edge_link() + 
  geom_node_point(aes(color = factor(color))) + 
  geom_node_text(aes(label = name), repel = TRUE) +
  theme_void() +
  theme(legend.position = "none") 

```

That is starting to look more meaningful!

```{r echo=TRUE}

# Set size to degree centrality 
V(grateful_data)$size = degree(grateful_data)

# Additional customisation for better legibility 
ggraph(grateful_data, layout = "fr") +
  geom_edge_arc(strength = 0.2, width = 0.5, alpha = 0.15) + 
  geom_node_point(aes(size = size, color = factor(color))) + 
  geom_node_text(aes(label = name, size = size), repel = TRUE) +
  theme_void() +
  theme(legend.position = "none") 

```

There is a lot more to do, but this is a great start.


*Citations:*

*Allan, Alex; Grateful Dead Lyric & Song Finder: https://whitegum.com/~acsa/intro.htm*

*ASCAP. 18 March 2022.*

*Dodd, David; The Annotated Grateful Dead Lyrics: http://artsites.ucsc.edu/gdead/agdl/*

*Schofield, Matt; The Grateful Dead Family Discography: http://www.deaddisc.com/*

*This information is intended for private research only, and not for any commercial use. Original Grateful Dead songs are ©copyright Ice Nine Music*



